### 2016-02-18 paper/paper.tex
Final abstract and paper revision.

\author{Aristotelis Spathis-Papadiotis\inst{1} \and Konstantinos Moustakas\inst{1}}
\authorrunning{Konstantinos Moustakas et al.} % abbreviated author list (for running head)

@@

\begin{abstract}
  This paper presents a simulation of a tsunami impact upon an urban coastline. Emphasis
  was given to the conservation of momentum during the simulation, as its distribution in
  space and time is the main factor of the wave's effects on the coastline. Due to this, a
  hybrid simulation method was adopted, based on the Smoothed Particle Hydrodynamics
  method, enriched with geometric constraints and rigid body interactions. The
  implementation is the result of cooperation between the Bullet physics engine and a
  custom SPH engine, which successively process the dynamic state of the fluid at every
  timestep of the simulation. Furthermore, in order to achieve better performance a custom
  data structure (LP grid) was developed for the optimization of locality in data storage
  and minimization of access time. Simulation data is exported to VTK files to allow
  interactive processing and visualization. \keywords{Fluid simulation, tsunami, SPH,
    tsunami-coastline interaction, force visualization}
\end{abstract}

@@

\subsection{SPH method}
Starting from the identity:
\begin{equation*}
  f(\vec{r}) = \int_Vf(\vec{x}) \delta(\vec{r} - \vec{x}) d\vec{x},
\end{equation*}
where $\delta(\vec{r})$ the Dirac delta function and $\vec{x} \in V$, one can obtain a
more general interpolation rule by substituting $\delta(\vec{r})$ with a smoothing kernel
$W(\vec{r}, h)$:
\begin{equation*}
  \label{eq:c-approx}
  f(\vec{r}) \approx \int_V f(\vec{x}) W(\vec{r}-\vec{x}, h) d\vec{x}
\end{equation*}
whose limit when $h\to0$ approaches the delta fucntion and is normalized to unity:
\begin{equation*}
  \label{eq:kernel-properties}
  \lim_{h\to0}W(\vec{r}, h) = \delta(\vec{r})\ \ \
  \text{and}\ \ \
  \int_VW(\vec{r}, h) d\vec{x} = 1
\end{equation*}
The smoothing radius $h$ serves as a cutoff radius in the smoothing process, as particles
beyond that distance have no contribution to the sum, i.e. $W(r,h) = 0$ when $r>h$. For
the discrete case, where $f$ is discretized to particles with density $\rho$ and mass $m$,
the weighting ratio $m/\rho$ can be used to construct a weighted sum interpolant for any
field $A$:
\begin{equation}
  \label{eq:d-approx}
  A(\vec{r}) = \sum_i \frac{m_i}{\rho_i} A(\vec{r}_i) W(\vec{r}-\vec{x}_i, h)
\end{equation}
which lies at the heart of SPH formulation. According to this, the gradient can be
computed by the following approximation:
\begin{equation}
  \label{eq:d-grad}
  \nabla A(\vec{r}) = \sum_i \frac{m_i}{\rho_i} A(\vec{r}_i) \nabla W(\vec{r} - \vec{x}_i, h)
\end{equation}
The obvious advantage of this is the exclusive dependence on the smoothing kernel
gradient, which can be precomputed for sensible kernel choices. However, this formula can
lead to unsymmetric pair forces, compromising the conservation of linear and angular
momentum of the system. To symmetrize these forces depending on gradients (like those
originating from pressure differences), we can use the product rule:
\begin{equation*}
  \label{eq:grad-identity}
  \nabla \left( \frac{P}{\rho} \right) =
  \frac{\nabla P}{\rho}-
  \frac{P}{\rho^2} \nabla \rho
  \hspace{10pt} \Leftrightarrow \hspace{10pt}
  \nabla P = \rho \left[ \frac{P}{\rho^2} \nabla \rho + \nabla \left( \frac{P}{\rho}
    \right) \right]
\end{equation*}
to obtain an alternative approximation of gradient
\begin{align}
  \label{eq:grad-est}
  \nabla P & = \rho \left[ \frac{P}{\rho^2} \sum_i \frac{m_i}{\rho_i} \rho \nabla
             W(\vec{r}-\vec{r}_i, h)
             +
             \sum_i \frac{m_i}{\rho_i} \frac{P_i}{\rho_i} \nabla W(\vec{r}-\vec{r}_i, h)
             \right] \nonumber \\
           & = \rho \sum_i m_i \left(\frac{P}{\rho^2} + \frac{P_i}{\rho_i} \right)
             \nabla W(\vec{r}-\vec{r}_i, h)
\end{align}
which is antisymmetric for all interacting particle pairs. Viscosity forces on the other
hand, are proportional to the laplacian of the velocity field:
\begin{equation}
  \label{eq:lapl-est}
  \nabla^2\vec{v} = \sum_i \frac{m_i}{\rho_i} (\vec{v}_i - \vec{v}) \nabla^2
  W(\vec{r}-\vec{r}_i, h)
\end{equation}
and are always antisymmetric, since they depend on velocity difference
$\vec{v}_i - \vec{v}$ between particles.  In each timestep of the simulation, the density
of all particles is first computed according to equation \ref{eq:d-grad}, as it depends
only on the relative position of those. The pressure at each particle location is then
obtained from its respective density through an equation of state. Subsequently, the
pressure and viscosity forces are computed from the particle data and integrated back into
the position and velocity of the particles.

@@

The Bullet physics engine is used to handle the rigid body dynamics. The simulation domain
is divided by a regular grid of spacing equal to the smoothing radius into cells
containing the fluid particles, which are generated as Bullet objects and stored in a
custom cell list data structure consisting of three vectors. As shown in figure
\ref{fig:lp-grid}, particles are accessed by following pointers through those vectors in
order, with the first encoding the 3D to linear locality preserving mapping of cells on
the second, and the second pointing at first of the cell's particles which are
continuously stored on the third, a dynamic sliding vector. This data structure allows for
quick neighbour search, interaction scanning, exploitation of access patterns of the SPH
algorithm and fast, in-place update. Simulation is following the Bullet framework, with
the SPH code being embedded as an internal timestep tick callback function. We took
advantage of the Bullet infrastructure to extract detailed information about the
collisions between fluid and terrain regarding the resulting impulse, time, and
location. Impulse and particle data are then written to multiple VTK files per
frame. Samples of the smoothed color field (common name in the literature for the field
having the value 1 at particle locations and 0 everywhere else) on the regular cell
lattice are also exported, which are then used to reconstruct the fluid surface as an
isosurface of that field. At the end of the simulation a cumulative impulse heatmap along
with the scaled and docked terrain model are also provided.

@@

Figure \ref{fig:visualization} shows a typical simulation result with the aid of
Paraview\textsuperscript{TM}, where the reconstructed wave surface and impulses exerted on
a selected region are visualized along the 3D terrain model and plotted on a histogram. A
comparison of the resulting impulse heatmap from the same tsunami impact upon an exposed
model and one with a protective seawall is shown in figure \ref{fig:impulse-field}. These
heatmaps confirm a well-documented observation in the literature, i.e. that tsunami energy
is absorbed mostly by the first obstacle in its way. This has been noted by Danielsen et
al. \cite{danielsen2005asian} and Kathiresan and Narayanasamy \cite{kathiresan2005601},
who both emphasized the protective role coastal vegetation (mangrove forests) played in
mitigating the impact effects of the 2004 Indian Ocean tsunami. The same conclusion has
been reached by Yanagisawa et al. \cite{yanagisawa200927} through theoretical
approximation of the phenomenon backed by relevant field data and measurements. Seawalls
have been constructed in high-risk regions as a common countermeasure against tsunami
hazards. Although the waves may be so large as to overtop such barriers, these cases are
somewhat rare, making seawalls an effective first line of defense.

>>>>

\author{Aristotelis Spathis-Papadiotis\inst{1} \and Konstantinos Moustakas\inst{1}}
\authorrunning{Aristotelis Spathis-Papadiotis et al.} % abbreviated author list (for running head)

@@

\begin{abstract}
  This paper presents a simulation of a tsunami impact upon an urban coastline. Emphasis
  was given to the conservation of momentum, as its distribution in space and time is the
  main factor of the wave's effects on the coastline. Due to this, a hybrid simulation
  method was adopted, based on the Smoothed Particle Hydrodynamics (SPH) method, enriched
  with geometric constraints and rigid body interactions. The implementation is the result
  of cooperation between the Bullet physics engine and our custom SPH engine, which
  successively process the dynamic state of the fluid at every timestep. Furthermore, in
  order to achieve better performance a custom data structure (LP grid) was developed for
  the optimization of locality in data storage and minimization of access time. Simulation
  data is exported to VTK files, allowing interactive processing and
  visualization. Experimental results demonstrate the benefits of impulse recording at
  potential hazard estimation and evaluation of defense strategies. \keywords{Fluid
    simulation, tsunami, SPH, tsunami-coastline interaction, force visualization}
\end{abstract}

@@

\subsection{SPH method}
Starting from the identity:
\begin{equation*}
  f(\vec{r}) = \int_Vf(\vec{x}) \delta(\vec{r} - \vec{x}) d\vec{x},
\end{equation*}
where $\delta(\vec{r})$ the Dirac delta function and $\vec{x} \in V$, one can obtain a
more general interpolation rule by substituting $\delta(\vec{r})$ with a smoothing kernel
$W(\vec{r}, h)$:
\begin{equation*}
  \label{eq:c-approx}
  f(\vec{r}) \approx \int_V f(\vec{x}) W(\vec{r}-\vec{x}, h) d\vec{x}
\end{equation*}
whose limit when $h\to0$ approaches the delta fucntion and is normalized to unity:
\begin{equation*}
  \label{eq:kernel-properties}
  \lim_{h\to0}W(\vec{r}, h) = \delta(\vec{r})\ \ \
  \text{and}\ \ \
  \int_VW(\vec{r}, h) d\vec{x} = 1
\end{equation*}
The smoothing radius $h$ serves as a cutoff radius in the smoothing process, as particles
beyond that distance have no contribution to the sum, i.e. $W(r,h) = 0$ when $r>h$. For
the discrete case, where $f$ is discretized to particles with density $\rho$ and mass $m$,
the weighting ratio $m/\rho$ can be used to construct a weighted sum interpolant for any
field $A$:
\begin{equation}
  \label{eq:d-approx}
  A(\vec{r}) = \sum_i \frac{m_i}{\rho_i} A(\vec{r}_i) W(\vec{r}-\vec{x}_i, h)
\end{equation}
which lies at the heart of SPH formulation. According to this, the gradient can be
computed by the following approximation:
\begin{equation}
  \label{eq:d-grad}
  \nabla A(\vec{r}) = \sum_i \frac{m_i}{\rho_i} A(\vec{r}_i) \nabla W(\vec{r} - \vec{x}_i, h)
\end{equation}
The obvious advantage of this is the exclusive dependence on the smoothing kernel
gradient, which can be precomputed for sensible kernel choices. However, this formula can
lead to unsymmetric pair forces, compromising the conservation of linear and angular
momentum of the system. To symmetrize these forces depending on gradients (like those
originating from pressure differences), we can use the product rule:
\begin{equation*}
  \label{eq:grad-identity}
  \nabla \left( \frac{P}{\rho} \right) =
  \frac{\nabla P}{\rho}-
  \frac{P}{\rho^2} \nabla \rho
  \hspace{10pt} \Leftrightarrow \hspace{10pt}
  \nabla P = \rho \left[ \frac{P}{\rho^2} \nabla \rho + \nabla \left( \frac{P}{\rho}
    \right) \right]
\end{equation*}
to obtain an alternative approximation of gradient
\begin{align}
  \label{eq:grad-est}
  \nabla P & = \rho \left[ \frac{P}{\rho^2} \sum_i \frac{m_i}{\rho_i} \rho \nabla
             W(\vec{r}-\vec{r}_i, h)
             +
             \sum_i \frac{m_i}{\rho_i} \frac{P_i}{\rho_i} \nabla W(\vec{r}-\vec{r}_i, h)
             \right] \nonumber \\
           & = \rho \sum_i m_i \left(\frac{P}{\rho^2} + \frac{P_i}{\rho_i} \right)
             \nabla W(\vec{r}-\vec{r}_i, h)
\end{align}
which is antisymmetric for all interacting particle pairs. Viscosity forces on the other
hand, are proportional to the laplacian of the velocity field:
\begin{equation}
  \label{eq:lapl-est}
  \nabla^2\vec{v} = \sum_i \frac{m_i}{\rho_i} (\vec{v}_i - \vec{v}) \nabla^2
  W(\vec{r}-\vec{r}_i, h)
\end{equation}
and are always antisymmetric, since they depend on velocity difference
$\vec{v}_i - \vec{v}$ between particles.  In each timestep of the simulation, the density
of all particles is first computed according to equation (\ref{eq:d-grad}), as it depends
only on the relative position of those. The pressure at each particle location is then
obtained from its respective density through an equation of state. Subsequently, the
pressure and viscosity forces are computed from the particle data and integrated back into
the position and velocity of the particles.

@@

The Bullet physics engine is used to handle the rigid body dynamics. The simulation domain
is divided by a regular grid of spacing equal to the smoothing radius into cells
containing the fluid particles, which are generated as Bullet objects and stored in a
custom cell list data structure consisting of three vectors. As shown in Figure
\ref{fig:lp-grid}, particles are accessed by following pointers through those vectors in
order, with the first encoding the 3D to linear locality preserving mapping of cells on
the second, and the second pointing at first of the cell's particles which are
continuously stored on the third, a dynamic sliding vector. This data structure allows for
quick neighbour search, interaction scanning, exploitation of access patterns of the SPH
algorithm and fast, in-place update. Simulation is following the Bullet framework, with
the SPH code being embedded as an internal timestep tick callback function. We took
advantage of the Bullet infrastructure to extract detailed information about the
collisions between fluid and terrain regarding the resulting impulse, time, and
location. Impulse and particle data are then written to multiple VTK files per
frame. Samples of the smoothed color field (common name in the literature for the field
having the value 1 at particle locations and 0 everywhere else) on the regular cell
lattice are also exported, which are then used to reconstruct the fluid surface as an
isosurface of that field. At the end of the simulation a cumulative impulse heatmap along
with the scaled and docked terrain model are also provided.

@@

Figure \ref{fig:visualization} shows a typical simulation result with the aid of
Paraview\textsuperscript{TM}, where the reconstructed wave surface and impulses exerted on
a selected region are visualized along the 3D terrain model and plotted on a histogram. A
comparison of the resulting impulse heatmap from the same tsunami impact upon an exposed
model and one with a protective seawall is shown in Figure \ref{fig:impulse-field}. These
heatmaps confirm a well-documented observation in the literature, i.e. that tsunami energy
is absorbed mostly by the first obstacle in its way. This has been noted by Danielsen et
al. \cite{danielsen2005asian} and Kathiresan and Narayanasamy \cite{kathiresan2005601},
who both emphasized the protective role coastal vegetation (mangrove forests) played in
mitigating the impact effects of the 2004 Indian Ocean tsunami. The same conclusion has
been reached by Yanagisawa et al. \cite{yanagisawa200927} through theoretical
approximation of the phenomenon backed by relevant field data and measurements. Seawalls
have been constructed in high-risk regions as a common countermeasure against tsunami
hazards. Although the waves may be so large as to overtop such barriers, these cases are
somewhat rare, making seawalls an effective first line of defense.

### 2016-02-18 report/abstract.tex
Final abstract and paper revision.

% \begin{minipage}[t][0.49\textheight][t]{\textwidth}
\section*{Περίληψη}
\paragraph{} Αντικείμενο της παρούσας εργασίας ήταν η ακριβής προσομοίωση της
πρόσ\-πτω\-σης ενός τσουνάμι σε ακτογραμμή. Ιδιαίτερη έμφαση δόθηκε στη διατήρηση της
ορμής, καθώς η κατανομή της στο χώρο και το χρόνο είναι καθοριστική για τις επιπτώσεις του
κύματος στην ακτογραμμή. Για το λόγο αυτό, υιοθετήθηκε μια υβριδική μέθοδος προσομοίωσης,
βασισμένη στη μέθοδο \eng{SPH} (\eng{Smoothed Particle Hydrodynamics}), εμπλουτισμένη όμως
με γεωμετρικούς περιορισμούς και αλληλεπιδράσεις υλικών σωμάτων. Η υλοποίηση βασίζεται στη
συνεργασία της μηχανής φυσικής \eng{Bullet} και της μηχανής \eng{SPH}, που επεξεργάζονται
αλληλοδιάδοχα την δυναμική κατάσταση του ρευστού σε κάθε βήμα της προσομοίωσης. Για την
καλύτερη απόδοση της προσομοίωσης αναπτύχθηκε επιπλέον ειδική δομή δεδομένων (\eng{LP
  grid}) για την βέλτιστοποίηση της τοπικότητας κατά την αποθήκευση των δεδομένων στη
μνήμη και της ταχύτητας πρόσβασης σε αυτά. Τα δεδομένα της προσομοίωσης εξάγονται σε
ειδικής μορφής αρχεία κειμένου (\eng{VTK}) με σκοπό την εκ των υστέρων διαδραστική
οπτικοποίηση και επεξεργασία μέσω κατάλληλων προγραμμάτων (όπως του \eng{ParaView}).

\paragraph{Λέξεις Κλειδιά} Προσομοίωση ρευστών, τσουνάμι, \eng{SPH}, αλληλεπίδραση
τσουνάμι-α\-κτο\-γραμ\-μής, οπτικοποίηση δυνάμεων
% \end{minipage}
\clearpage
% \begin{minipage}[b][0.49\textheight][t]{\textwidth}
\selectlanguage{english}
\section*{Abstract}
\paragraph{} The objective of this thesis was the accurate simulation of a tsunami impact
upon a coastline. Emphasis was given to the conservation of momentum during the
simulation, as its distribution in space and time is the main factor of the wave's effects
on the coastline. Due to this, a hybrid simulation method was adopted, based on the SPH
method (Smoothed Particle Hydrodynamics), enriched with geometric constraints and rigid
body interactions. The implementation is the result of cooperation between the Bullet
physics engine and a custom SPH engine, which successively process the dynamic state of
the fluid at every timestep of the simulation. Furthermore, in order to achieve better
performance a custom data structure (LP grid) was developed for the optimization of
locality in data storage and minimization of access time. Simulation data is exported to
text files in VTK format to allow interactive processing and visualization with the aid of
specialized programs (like ParaView).
\paragraph{Keywords} Fluid simulation, tsunami, SPH, tsunami-coastline interaction,
force visualization

>>>>

% \begin{minipage}[t][0.49\textheight][t]{\textwidth}
\section*{Περίληψη}
\paragraph{} Αντικείμενο της παρούσας εργασίας ήταν η προσομοίωση της πρόσ\-πτω\-σης ενός
τσουνάμι σε ακτογραμμή πόλης. Έμφαση δόθηκε στη διατήρηση της ορμής, καθώς η κατανομή της
στο χώρο και το χρόνο είναι καθοριστική για τις επιπτώσεις του κύματος στην
ακτογραμμή. Για το λόγο αυτό, υιοθετήθηκε μια υβριδική μέθοδος προσομοίωσης, βασισμένη στη
μέθοδο \eng{Smoothed Particle Hydrodynamics} (\eng{SPH}), εμπλουτισμένη όμως με
γεωμετρικούς περιορισμούς και αλληλεπιδράσεις υλικών σωμάτων. Η υλοποίηση είναι αποτέλεσμα
της συνεργασίας της μηχανής φυσικής \eng{Bullet} και μιας μηχανής \eng{SPH}, που
επεξεργάζονται αλληλοδιάδοχα την δυναμική κατάσταση του ρευστού σε κάθε χρονικό
βήμα. Επιπλέον, για την καλύτερη απόδοση της προσομοίωσης αναπτύχθηκε ειδική δομή
δεδομένων (\eng{LP grid}) για τη διατήρηση της τοπικότητας κατά την αποθήκευση των
δεδομένων στη μνήμη και της ελαχιστοποίησης του χρόνου πρόσβασης σε αυτά. Τα δεδομένα της
προσομοίωσης εξάγονται σε ειδικής μορφής αρχεία κειμένου (\eng{VTK}), επιτρέποντας την εκ
των υστέρων διαδραστική οπτικοποίηση και επεξεργασία. Πειραματικά αποτελέσματα δείχνουν τα
πλεονεκτήματα της καταγραφής ώσεων για εκτίμηση πιθανής καταστροφής και την αξιολόγηση
στρατηγικών προστασίας.

\paragraph{Λέξεις Κλειδιά} Προσομοίωση ρευστών, τσουνάμι, \eng{SPH}, αλληλεπίδραση
τσουνάμι-α\-κτο\-γραμ\-μής, οπτικοποίηση δυνάμεων
% \end{minipage}
\clearpage
% \begin{minipage}[b][0.49\textheight][t]{\textwidth}
\selectlanguage{english}
\section*{Abstract}
\paragraph{} The objective of this thesis was the simulation of a tsunami impact upon an
urban coastline. Emphasis was given to the conservation of momentum, as its distribution
in space and time is the main factor of the wave's effects on the coastline. Due to this,
a hybrid simulation method was adopted, based on the Smoothed Particle Hydrodynamics (SPH)
method, enriched with geometric constraints and rigid body interactions. The
implementation is the result of cooperation between the Bullet physics engine and our
custom SPH engine, which successively process the dynamic state of the fluid at every
timestep. Furthermore, in order to achieve better performance a custom data structure (LP
grid) was developed for the optimization of locality in data storage and minimization of
access time. Simulation data is exported to VTK files, allowing interactive processing and
visualization. Experimental results demonstrate the benefits of impulse recording at
potential hazard estimation and evaluation of defense strategies.

\paragraph{Keywords} Fluid simulation, tsunami, SPH, tsunami-coastline interaction, force
visualization
